% !TeX spellcheck = en_GB
\chapter{Environments}
\label{chap:environment}
In this chapter we will discuss the environments we have chosen to use in our examination of evolutionary algorithms and the theoretical background that will help us analyse them.

\section{Reinforcement learning}
\label{sec:RL}
Reinforcement learning might be best introduced by description of the problem it aims to solve.
\begin{defn}[Markov’s decision problem]
Let P be a probability measure over the states of the environment. We define Markov’s decision problem (MDP) by a four-tuple (S, A, T, r) where:
\begin{itemize}
	\item S is a set of all possible states of the system, or state space
	\item A is a set of all possible actions available in the system or action space
	\item T is a transition function:
	$$\forall s',s \in S, a\in A: T(s, a,  s')=P(s'|s,a)$$
	\item $ r: S\times A\times S \rightarrow \boldsymbol{R} $ is reward function, $ r(s, a, s') ;$ being the reward received when the execution of action a in state s leads to state s'.
\end{itemize}
\end{defn}
To tie this into the information from chapter \ref{ch:agents} $ A=A(E)$, $ R $ is performance measure of an agent solving the MDP and the percepts have bijection to real states of the environment. This would not be the case for partially observable MDP.
\begin{defn}[\cite{KAELBLING199899}]
Let P be a probability measure over the states of the environment.	
We define Partialy Observable Markov’s decision problem (POMDP) by a four-tuple (S, A, $\Omega$, T, O, r) where:
\begin{itemize}
	\item S is a set of all possible (true) states of the system, or state space
	\item A is a set of all possible actions available in the system or action space
	\item $\Omega$ is a set of all possible observations on the system, or observation space
	\item $T: S\times A\times S \rightarrow [0, 1]$: is a transition function over true states. 
	$$\forall s',s \in S, \forall a\in A: T(s, a,  s')=P(s'|s,a)$$.
	
	\item $ O: S\times A \times \Omega \rightarrow [0, 1]$. $$\forall s' \in S, \forall a\in A, \forall o\in\Omega: O(s', a,  o)=P(o|a, s')$$ or the probability of making observation o given that the agent took action a and landed in state s’. 
	\item $ r: S\times A\times S \rightarrow \boldsymbol{R} $ is reward function same as in MDP.
\end{itemize}
\end{defn}
Now that we have characterised the environments we can apply to them concepts we have revied in the chapter \ref{ch:agents}. We begin by specifying the notion of percept sequence space from definition \ref{dfn:actuators}.
\begin{defn}[\cite{DBLP:journals/corr/RossPPC14}]
	Let $ \Omega $ the space of all observations ($ \Omega \cong S $ for MDP), A be the action space, $ o_t \in \Omega $ be observation at time $ t $ and $ a_t \in A $ action taken by the agent in time $ t $. Then $$ h_t = (a_0, o_1,...,a_{t-1},o_t) $$ is a history at time $ t $.
\end{defn}
\begin{defn}[\cite{DBLP:journals/corr/RossPPC14}]
	Let P be probability measure over states of the environment S and $ s_t \in S $ be the state of the environment at time $ t $ . Then function $ b_t: S \rightarrow [0,1]$ such that $$ b_t(s) = P(s_t = s|h_t, b_0)$$ is called a belief state of an agent, where $ b_0 $ is the apriori belief state and $ h_t $ is history at time $ t $. 
\end{defn}
We also need to reframe the agent function discussed in definition \ref{dfn:actuators}.
\begin{defn}[\cite{DBLP:journals/corr/RossPPC14}]
	Let P be a probability measure and A be an action space. Then function $ \pi $ defined as  $ \pi(b_t, a)=P(a|b_t)$ where $ P(a|b_t) $ denotes probability of doing action $ a $ given belief state $ b_t $ is called agent policy. 
\end{defn}
\begin{defn}[\cite{DBLP:journals/corr/RossPPC14}]
	\label{def:optpoc}
	Let r be a reward function, A an action space, T a transition function, $\gamma \in [0,1) $ a discount factor and $ \Pi $ be a set of all possible agent policies, then
\[
\pi^* =
\arg\max_{\pi \in \Pi}
\mathbb{E}
\left[
\sum_{t=0}^{\infty}
\gamma^t
\sum_{s \in S}
b_t(s)
\sum_{a \in A}\sum_{s' \in S}
r(s,a, s')T(s,a,s')\pi(b_t,a)
\;\middle|\;
b_0
\right]
\] is called optimal policy.
\end{defn}
The goal of reinforcement learning is then to proximate optimal policy by trial and error, based on the received reward, as accurately as possible. 
This can be done utilising various online or offline methods such as Q-learning. 
In the literature  and specifically in the cited \cite{DBLP:journals/corr/RossPPC14} the the definition \ref{def:optpoc} features $ \gamma $ as an discounting factor for the reward. This is a crucial parameter for reinforcement learning algorithms that require continuous rewards. However as we will discuss further in this chapter and chapter \ref{chap:evolution}, in evolutionary algorithms we evaluate the policy as a whole and are only presented with final value calculated. Therefore it is not necessary to know or discover particular value of $ \gamma $ for the environments presented in this thesis. 

Note that since in this thesis we are only concerned with simple reflexive agents, the belief state is not explicitly calculated statistic but is nonetheless reflected in the learned reactions to observation.
 
\section{Gymnasium}
We intend to set our agents into environments provided by Gymnasium project which emerged from it's predecessor OpenAI Gym project. Both of them Python libraries, trying to provide researches with benchmark reinforcement learning problems in form of simple games, sometimes even recreating old console games such as Air Raid or Backgammon.

While Gymnasium provides wide range of environments, quite a number of them if not most do not provide their outputs in structured form. Instead the resulting screen has to be processed by user defined visual processor to extract information crucial for completing the task. 

This introduces a layer of complexity that is outside of the scope of this thesis. Therefore we have decided to narrow our selection only to environments which do provide structured output. We describe the loop of receiving and providing feedback between agent and environment in algorithm \ref{alg:envloop}.
\begin{algorithm}[h]
	\caption{Environment Interaction Loop}
	\label{alg:envloop}
	
	\KwIn{Environment $\mathcal{E}$}
	\KwIn{Number of episodes $ n $}
	\KwIn{Observation aggregation functions $agg$ and $ transform $}
	R = 0\;
	\For{i =0 \KwTo n}{
	$o \leftarrow \mathrm{reset}(\mathcal{E})$\;
	$O \leftarrow \mathrm{reset}(O)$\;
	\While{not terminated}{
		
		$a \leftarrow \mathrm{Agent}(o)$\;
		
		$(o, r, terminated)
		\leftarrow
		\mathrm{Environment}(a)$\;
		$ R \leftarrow R + r $\;
		$ O  \leftarrow agg(O, o)$\;
	}
	}
	\Return R/n, $ transform(O) $;
	
\end{algorithm}


 As mentioned in the introduction the selected environments are called Cartpole and Lunar Lander. 
 
 We will discuss the full purpose of $ agg $ and $ transform $ in the section \ref{sec:novelty} and section \ref{sec:behavioralSpace}.

\subsection{Cartpole}
\label{env:cartpole}

Cartpole is one of the most used benchmark environments in the Gymnasium library. It's a simple environment where the agent controls a cart rolling on flat surface with a pole on top.
The goal is to keep the pole balanced on top of the cart. The reward is calculated every step depending on the angle of the stick. This means we have bounded total reward between 0 and 500 points.
\begin{defn}
	An agent that achieves score 500 in Cartpole is called solution or solving individual.
\end{defn}
\subsubsection{Observation space}
Cartpole observation space is expressed as a 4-dimensional vector. We delineate possible values of the vector in \ref{tab:cartpole_state_space}. Note that for Cartpole enviroment this observation space reflects full state of the environment in the given moment and therefore it could be characterised as MDP rather than POMDP.
\begin{table}[h]
	\centering
	\begin{tabular}{|c|l|c|c|}
		\hline
		\textbf{Index} & \textbf{State Variable} & \textbf{Min} & \textbf{Max} \\
		\hline
		0 & Cart position & -4.8 & 4.8 \\
		1 & Cart velocity & $-\infty$ & $\infty$ \\
		2 & Pole angle & $-0.418$ rad & $0.418$ rad \\
		3 & Pole angular velocity & $-\infty$ & $\infty$ \\
		\hline
	\end{tabular}
	\caption{Observation space of Cartpole}
	\label{tab:cartpole_state_space}
\end{table}
\subsubsection{Action space}
The action space is discrete with only two values possible as presented in the table \ref{tab:cartpole_actions}. Note that we cannot chose to do nothing. 
\begin{table}[h]
	\centering
	\begin{tabular}{|c|l|}
		\hline
		\textbf{Action} & \textbf{Description} \\
		\hline
		0 & Push cart to the left \\
		1 & Push cart to the right \\
		\hline
	\end{tabular}
	\caption{Cartpole Action Space.}
	\label{tab:cartpole_actions}
\end{table}
\subsection{Lunar Lander}
The other environment we have selected is Lunar Lander where the agent controls descend of a landing shuttle on rugged planetary (or the Moon's) surface with the help of oddly placed rockets. The points are subtracted for staying in the air or for getting the shuttle into a spin. Bonus points awarded for landing in the centre of the  The shuttle has infinite fuel so the game ends only when the shuttle lands safely or crushes into the ground. Luckily we can at least limit the number of steps for which the environment is allowed to go on. For a neural network training this however poses a specific kind of problem. Firstly very narrow behavioural window that can achieve positive total reward. But mainly as opposed to the Cartpole environment (\ref{env:cartpole}), now the unsuccessful runs may take significantly longer time than the successful ones. This means experiments are particularly costly for slowly converging algorithms. The upper limit of sum total reward is 200, the lower would possibly be $-\inf$, depending on the length limitation imposed. 
We therefore differentiate between several stages of obtaining successful agent.
\begin{defn}
	An agent that achieves score 200 in Lunar lander is called solution or solving individual.
\end{defn}
\begin{defn}
	\label{def:wellbehaved}
	An agent that achieves positive score  in Lunar lander is called well-behaved.
\end{defn}
\begin{defn}
	\label{def:verywellbehaved}
	An agent that achieves average score over 100 in Lunar lander is called very well-behaved.
\end{defn}
 Definition \ref{def:wellbehaved} stems from the fact that it can only be achieved if the agent does not crash the shuttle as doing so yields a penalisation of -100 points while not obtaining a boost of +100 points for safe landing, driving even a possible solution with score 200 to 0. Similairly for definition \ref{def:verywellbehaved}, it is highly unlikely to average score over 100 without at least once landing with perfect score 200.
\subsubsection{Observation space}
The observation space of Lunar Lander is represented as  8 dimensional vector. Delineation of its values can be found in the table \ref{tab:lander_state_space}. Note that while the observation space is quite descriptive it does not capture all the information about environment’s state directly such as wind velocity or proximity to the ground. Because the landscape (except for a landing pad in the centre) might be different in each episode, we can only infer their parameters from the observations (such as velocity and leg contact with ground) indirectly. For simplicity we do not allow wind in our episodes as is the default setting of the environment. This means that the distinction between observation and true state becomes most tangible in low altitudes where agent risks either collision with unexpected hill or infinite hovering above unperceived valley, forgoing the +100 point reward. 
\begin{table}[h]
	\centering
	\begin{tabular}{|c|l|c|c|}
		\hline
		\textbf{Index} & \textbf{State Variable} & \textbf{Min} & \textbf{Max} \\
		\hline
		0 & $x$ position & -2.5 & 2.5 \\
		1 & $y$ position & -2.5 & 2.5 \\
		2 & $v_x$ velocity in x-axis & -10 & 10 \\
		3 & $v_y$ velocity in y-axis & -10 & 10 \\
		4 & angle $\theta$ & -6.2831855 & 6.2831855 \\
		5 & angular velocity & -10 & 10 \\
		6 & left leg contact (boolean) & 0 & 1 \\
		7 & right leg contact (boolean) & 0 & 1 \\
		\hline
	\end{tabular}
	\caption{Observation space of Lunar Lander}
	\label{tab:lander_state_space}
\end{table}
\subsubsection{Action space}
Action space for Lunar Lander can come in two shapes. First is discrete where we the agent chooses a number representing one of the engines to ignite. Second is continuous where agent has continuous control over the ignition of the main engine and one of the side engines. This variant allows for more precision during the flight but at the same time makes the policy the agent needs to learn significantly more complex, requiring more resources during the learning phase. To preserve compatibility with Cartpole setup and considering feasibility of multiple experiments ahead we will work with the discrete version of the action space outlined in table \ref{tab:llactionspace}.\\
.
\begin{table}[h]
	\centering
	\begin{tabular}{|c|l|}
		\hline
		\textbf{Action} & \textbf{Description} \\
		\hline
		0 & Do nothing \\
		1 & Fire left orientation engine \\
		2 & Fire main engine \\
		3 & Fire right orientation engine \\
		\hline
	\end{tabular}
	\caption{Action space definition for the control system.}
	\label{tab:llactionspace}
\end{table}




 
 
