\chapter*{Conclusion}
\addcontentsline{toc}{chapter}{Conclusion}
%Co práce udělala.
%Co bylo zjištěno. +
%Co je hlavní přínos. +
%Co nevyšlo / omezení.
%Co by měla dělat navazující práce.


We have successfully developed and implemented framework for testing evolutionary strategies and differential evolution in several setups, such as fitness, novelty and their respective archiving and a combination of them. We have also proposed behavioural characterisations for both environments Lunar Lander and CartPole. 
We have used the characteristics to investigate performance of objective handling methods derived from fitness, novelty and their combinations. We used the same series of experiments to determine a baseline hyper-parameter setting for each combination of environment, EA and objective handling method. On these setting we have performed statistical tests to determine relationships between the objective handling methods and their groupings based on the main objective used (fitness, novelty or combination).    
The areas of comparisons were: absolute performance, rate of convergence and diversity of the final populations.

We have identified that in the selected environments the strongest performance was obtained using the fitness based methods followed by the combinations with novelty. Novelty based methods were the weakest performing. We link this outcome to the complexity of selected environments. The CartPole environment is so simple that we obtain the solving individual through fitness only, the novelty can only provide more diverse populations. While Lunar Lander clearly has false optima the fitness based methods fall into, only a small strain of behaviours can reach it or surpass it. The novelty search attempting to explore the whole behavioural space is therefore rendered inefficient. 
We have also confirmed that dynamics of the search are very important. The elitist version of novelty archive with static threshold could generally not outperform random novelty archive in clear antipattern to the fitness archives while either dynamic combinations of novelty were generally able to outperform both novelty archives. 
The results suggest that approach combining novelty and fitness or at least introducing controlled diversity into the population is therefore necessary to reliably obtain full solution. This confirms assumptions in literature delineating the usefulness of novelty such as  \cite{noveltyarticle} or \cite{notenoughnovelty}.

Main contribution of this thesis is the empirical evaluation of novelty-based objective handling methods in evolutionary reinforcement learning. The work provides a systematic comparison of fitness-based selection, novelty search, their archival variants, and a dynamic combination of fitness and novelty on two low deception benchmark environments. In addition, the thesis analyses the impact of these methods on optimisation performance and population diversity. It also proposes baseline behavioural characteristics for the selected environments and discusses the effects of the design on the analysis.  


Our analysis was limited by the  choice of behavioural characteristics being driven mostly by interpretability and not raw performance. Specifically for Lunar Lander it is possible that a more performance focused choice, perhaps utilising automated extraction of behavioural characteristics \citep{SDBC} would provide a clearer picture. However optimisation of behavioural space was out of the scope of this thesis.
Other limitation is the choice of evaluation scope for the Lunar Lander where we limited the number of steps and the action space. 

In continuous version of Lunar Lander or given more steps the agents might have been able to fines the last stage of landing more appropriately even without the use of novelty. Loosening of these limitations however has tangible impact on effective performance of the evolutionary search as a whole. 
We have also chose to limit dynamics of the evolutionary strategy algorithm.

Several directions for future work emerge from this thesis. First the experiments could be extended to a broader range of reinforcement learning environments to determine how broadly can our findings be generalised. This includes investigation of multiagent environments where the reward is based on behaviour of more than one agent such as the environment Waterworld from the Gymnasium library. Secondly an investigation and cross analysis of the automatic algorithms for behavioural space extraction on the environments. Last but not least a broader selection of evolutionary algorithms could be investigated as well, with focus on dynamic properties of the algorithms. 

%Despite, the general approach shows greater promise as we could expect that a more powerful behavioural characteristic might be discovered using automatic methods that were outside of the scope of this thesis. On the other hand there is a limited room to refine the fitness based archive substantially over the version presented in this text without altering its core concept.  %>



%The occasional dominance of increasing fitness over increasing novelty also seems conditioned on the structure of the environments with Cartpole rewarding similar behaviour consistently throughout it's episodes  and Lunar Lander having only behavioural bottleneck at the later stage of the episode.%>

%Originally we have also investigated Gymnasium's environment Waterworld where the fitness structure depends on other agents evolving. Ultimately it was deemed out of scope for this thesis but it is an interesting direction of these investigations left for the future works.


